A parametricity-based formalization of semi-simplicial and semi-cubical sets

### 2023-05-28 paper/paper.tex
paper: polish first few sections; removed some undef refs

\subsection*{Fibered vs indexed presentation of semi-simplicial and semi-cubical sets}

A family of sets can commonly be represented in two ways: as a family properly speaking, indexed over a set or as a set together with a map specifying for each element of the set on which index it depends. In the latter case, the set associated to a given index is the fiber, i.e. inverse image, of the index under the map. The two presentations are equivalent, what can be formally stated in a type-theoretic language as a fibered/indexed correspondence~\footnote{At a discussion on the Zulip channel for the HoTT community, alternative names fibration/family correspondence and unbundled/bundled correspondence were proposed, among others. In our experience, the opposition fibered/indexed is spontaneously understood by many listeners. In particular, it echoes to the Grothendieck construction of fibered categories from indexed categories. The most elementary instance of the correspondence, with $\Type$ instead of $\U$, that is $(\Sigma S: \Type. (S \rightarrow T)) \simeq (T \rightarrow \Type)$, is sometimes called ``Grothendieck construction for dummies'' (its proof requires univalence, see~\cite{hottbook}).}:
\begin{equation*}
  \mbox{(fibered)}\qquad(\Sigma S: \U. (S \rightarrow T)) \simeq (T \rightarrow \U)\qquad \mbox{(indexed)}
\end{equation*}
where $\U$ represents the subset of types with a unique proof of equality, that is the type-theoretic counterpart of the category of sets (within some universe).

@@

\begin{align*}
  X_0 & :              & \U                                                     \\
  X_1 & :              & X_0 \rightarrow X_0 \rightarrow \U                     \\
  X_2 & : \Pi a b c d. & X_1(a)(b) \rightarrow X_1(a)(c) \rightarrow X_1 (b)(d) \\
      &                & \rightarrow X_1 (c)(d) \rightarrow \U                  \\
  \ldots
\end{align*}
%% \begin{align*}
%%   X_0                                                                                      & \mbox{ a set of points}                                       \\
%%   (X_1)_{a,b\in X_0}                                                                       & \mbox{ a set of lines between any two points}                 \\
%%   (X_2)_{a,b,c,d\in X_0,p\in (X_1)_{ab},q\in (X_1)_{bc},r \in (X_1)_{ac}, s\in (X_1)_{cd}} & \mbox{ a set of squares over any 4 lines lying over 4 points} \\
%%   \ldots
%% \end{align*}

The idea for such indexed presentation of presheaves over a direct category was introduced at the Univalent Foundations year in the context of defining semi-simplicial types~\cite{}. A couple of constructions were proposed since then. One, by \cite{voevodsky12}, relies on the presentation of semi-simplicial sets as a presheaf over increasing injective maps between finite ordinals. Another, by \cite{herbelin15}\footnote{The title of the paper ``A dependently-typed construction of semi-simplicial types'' is confusing: it was part of a project at defining semi-simplicial types, but, in practice, the definition is done in a type theory with Unicity of Identity Proofs (UIP), meaning in a type theory where types (intended to represent arbitrary $\infty$-groupoids) are actually sets ($0$-truncated $\infty$-groupoids, that is, morally, $0$-groupoids).}, and fully formalized in Coq, relies on the presentation of semi-simplicial sets as a presheaf over face maps. Two other ones by~\cite{part15}, formalized in (an emulation of logic-enriched homotopoty type theory in) the Plastic proof assistant, and by \cite{annenkovCK17}, formalized in (an emulation of a two-level type theory in) the Agda proof assistant~\footnote{\url{https://github.com/nicolaikraus/HoTT-Agda/blob/master/nicolai/SemiSimp/SStypes.agda}}, also rely on the presentation of the semi-simplicial category from increasing injective func†ions over finite ordinals.

The indexed definition of a presheaf over a direct category is technically more involved than the presheaf definition as it requires hard-wiring in the structure the dependencies between elements of the sets of the presheaf, including the coherence conditions between these dependencies, such as taking the $i$-th face of the $j$-th face of a $n$-simplex is the same as taking the $(j-1)$-th face of the $i$-th face (when $j>i$). On the other side, defining a concrete presheaf in indexed form only requires to provide the family since the burden of defining maps and showing the coherence conditions has already been integrated in the definition what is not the case in the presheaf presentation.

\subsection*{Reynolds' parametricity and its unary and binary variants}
In the context of functional programming, Reynolds' parametricity~\cite{reynolds83} interprets types as relations characterizing the observational behavior of programs of this type. Parametricity can be iterated, and it has been noted that iterated Reynolds' parametricity has the structure of a cubical set~\cite{johann17,altenkirch15,moulin16,moeneclaey21,moeneclaey22phd}. We obtain a \emph{unary} variant of Reynolds' \emph{binary} parametricity by using predicates or families instead of relations or graphs, in which case, we obtain a form of realizability~\cite{bernardy12,lasson12,moulin16}. It has then been noted that iterated unary parametricity has the structure of an augmented simplicial set~\footnote{Private communication with Hugo Moeneclaey and Thorsten Altenkirch}.

\subsection*{Iterated-parametricity-based construction}
We exploit the connection between iterated unary parametricity and augmented simplicial sets, and between iterated binary parametricity and cubical sets to present a uniform construction of both augmented semi-simplicial and semi-cubical sets, generated by iterating the parametricity translation for type judgements. In contrast to the standard presheaf presentation of augmented semi-simplicial and semi-cubical sets~\cite{fri08,grandis03,buchholtz17}, our uniform construction, which we call \emph{$\nu$-sets}, is \emph{indexed}. That is, instead of having the set of augmented semi-simplices or semi-cubes in dimension $n+1$ \emph{fibered} over the set of augmented semi-simplices or semi-cubes in dimension $n$, we consider families of augmented semi-simplices and semi-cubes indexed over their faces.

\subsection*{Contribution}
As a result, our work can be seen as contributing the following: it characterizes, will full details, how to technically define parametricity iterated in arbitrary dimensions; it provides a new example of indexed constructions of semi-simplicial and semi-cubical sets~\cite{voevodsky12,herbelin15,part15}, useful in the direction of better understanding the technicality of coherence issues in defining ``semi-simplicial types''~\cite{part15,shulman14,altenkirch16,kraus2017space}, as well as the dialectic between reasoning in Extensional Type Theory and Intensional Type Theory (as will be discussed in Section~\ref{sec:ett}); it makes a step in the direction of the program initiated in~\cite{altenkirch15} to develop parametricity-based models of parametric type theory~\cite{bernardy15,nuyts17,cavallo19} and cubical type theory~\cite{bezem13,cohen16,angiuli21} which are closer to the syntax of type theory and thus more liable to reflect definitional properties of type theory than presheaf-based cubical sets would do (compare e.g. to the loss of definitional properties when interpreting ``indexed'' dependent types of type theory as ``fibrations'' in models such as locally cartesian closed categories~\cite{curien14}).

\subsection*{Outline}

The outline of the paper is as follows. We recall the definition of the augmented semi-simplicial and semi-cubical categories using a combinatorial presentation, that directly leads to our generalization to $\nu$-sets, in \ref{sec:nu}. We then proceed to explain the connection between the indexed representation and parametricity in \ref{sec:indexed-param}. Before spelling out the details of our formalization in type-theoretic language in \ref{sec:ett} and \ref{sec:itt}, we give intuitions in \ref{sec:intuition}. We end with some of the finer details of our mechanization in \ref{sec:mech}.

@@

\section{Semi-simplicial and semi-cubical sets\label{sec:nu}}
In this section, we generalize semi-simplicial and semi-cubical sets to, what we call, $\nu$-sets, subsuming the earlier definitions. However, we also explicitly work out the cases of semi-simplicial and semi-cubical sets, in a way that is similar to a previous art~\cite{aitchison86}.

@@

\section{Preliminaries}
Martin-Löf's Type theory~\cite{martinlof75,martinlof84} is a logical formalism based on the notion of a \emph{type} rather than that of a \emph{set}. It can be seen as a foundation of mathematics alternative to set theory and is the core of several tools for the formalization of mathematics such as Agda~\cite{agda23}, Coq~\cite{coq22} or Lean~\cite{lean15}. In type theory, propositions are types and proofs are programs. A particularity of type theory is also that types and programs, hence propositions and proofs also, are considered modulo an equational theory called definitional equality.

@@

Tables \ref{fulltab:coind}, \ref{fulltab:core}, and \ref{fulltab:frames} are the same as tables \ref{tab:coind}, \ref{tab:core}, and \ref{tab:frames}, except that, in function applications, so as to be more informative, we make explicit all arguments, including those that can be inferred from the context.

The need for transport along a proof of commutation of $\restrf$ in the definition of $\restrl$ is made explicit as shown in table~\ref{fulltab:faces}, where the arrow over $\cohf$ indicates the direction of rewrite. The proof of $\cohf$ is described on~\ref{fulltab:coh}, and it requires making explicit several rewritings which were invisible in extensional type theory. The commutation of $\restrl$ lives in a type referring to $\cohf$, so we need a transport along the commutation of $\restrf$ in the statement of $\cohl$. The proof of $\cohl$ is the most involved proof in the construction. This is where the higher-dimensional coherence condition is needed. The exact formulation of this coherence condition is as follows.
\renewcommandx{\restrf}[7][1,2,3,4,5,6,7]{\restr{frame}[][#2][#3][#4][#5][#6][#7]}
\renewcommandx{\cohf}[9][1,2,3,4,5,6,7,8,9]{\coh{frame}[][#2][#3][#4][#5][#6,#7][#8][#9]}
\renewcommandx{\cohl}[9][1,2,3,4,5,6,7,8,9]{\coh{layer}[][#2][#3][#4][#5][#6,#7][#8][#9]}
\renewcommandx{\cohc}[9][1,2,3,4,5,6,7,8,9]{\coh{painting}[][#2][#3][#4][#5][#6,#7][#8][#9]}
\begin{align*}
  \cohf[m][\omega][\theta][r][p][n][p][][d = {\restrf[m][\epsilon][q+2][n][p][][d = \d]}]\;\bullet     \\
  \ap \restrf[m][\omega][r][n][p][][]\;(\cohf[m][\epsilon][\theta][q+1][p][n][p][][d = \d])\;\bullet   \\
  \cohf[m][\epsilon][\omega][q][r][n][p][][d = {\restrf[m][\theta][p][n][p][][d = \d]}] =              \\
  \ap \restrf[m][\theta][p][n][p][][]\;(\cohf[m][\epsilon][\omega][q+1][r+1][n][p][][d = \d])\;\bullet \\
  \cohf[m][\epsilon][\theta][q][p][n][p][][d = {\restrf[m][\omega][r+1][n][p][][d = \d]}]\;\bullet     \\
  \ap \restrf[m][\epsilon][q][n][p][]\;(\cohf[m][\omega][\theta][r][p][n][p][][d = \d])
\end{align*}
where $\mathsf{ap}$ applies a function on two sides of an equality, and $\bullet$ is transitivity of equality. This property of equality proofs holds in \U, and since our formalization is done in \U, the term is trivially discharged.

>>>>

\subsection*{Fibered vs indexed presentation of semi-simplicial and semi-cubical sets}
A family of sets can commonly be represented in two ways: as a family properly speaking, indexed over a set or as a set together with a map specifying for each element of the set on which index it depends. In the latter case, the set associated to a given index is the fiber, i.e. inverse image, of the index under the map. The two presentations are equivalent, what can be formally stated in a type-theoretic language as a fibered/indexed correspondence~\footnote{At a discussion on the Zulip channel for the HoTT community, alternative names fibration/family correspondence and unbundled/bundled correspondence were proposed, among others. In our experience, the opposition fibered/indexed is spontaneously understood by many listeners. In particular, it echoes to the Grothendieck construction of fibered categories from indexed categories. The most elementary instance of the correspondence, with $\Type$ instead of $\U$, that is $(\Sigma S: \Type. (S \rightarrow T)) \simeq (T \rightarrow \Type)$, is sometimes called ``Grothendieck construction for dummies'' (its proof requires univalence, see~\cite{hottbook}).}:
\begin{equation*}
  \mbox{(fibered)}\qquad(\Sigma S: \U. (S \rightarrow T)) \simeq (T \rightarrow \U)\qquad \mbox{(indexed)}
\end{equation*}
where $\U$ represents the subset of types with a unique proof of equality, that is the type-theoretic counterpart of the category of sets (within some universe).

@@

\begin{align*}
  X_0 & :              & \U                                                     \\
  X_1 & :              & X_0 \rightarrow X_0 \rightarrow \U                     \\
  X_2 & : \Pi a b c d. & X_1(a)(b) \rightarrow X_1(a)(c) \rightarrow X_1 (b)(d) \\
      &                & \rightarrow X_1 (c)(d) \rightarrow \U                  \\
  \ldots
\end{align*}

The idea for such indexed presentation of presheaves over a direct category was mentioned at the Univalent Foundations year in the context of defining semi-simplicial types~\footnote{See e.g. \url{https://ncatlab.org/nlab/show/semi-simplicial+types+in+homotopy+type+theory}.}. A couple of constructions were proposed since then. One, by \cite{voevodsky12}, relies on the presentation of semi-simplicial sets as a presheaf over increasing injective maps between finite ordinals. Another, by \cite{herbelin15}\footnote{The title of the paper ``A dependently-typed construction of semi-simplicial types'' is confusing: it was part of a project at defining semi-simplicial types, but, in practice, the definition is done in a type theory with Unicity of Identity Proofs (UIP), meaning in a type theory where types (intended to represent arbitrary $\infty$-groupoids) are actually sets ($0$-truncated $\infty$-groupoids, that is, morally, $0$-groupoids).}, and fully formalized in Coq, relies on the presentation of semi-simplicial sets as a presheaf over face maps. Another by~\cite{part15}, formalized in (an emulation of logic-enriched homotopy type theory in) the Plastic proof assistant, and yet another by \cite{annenkovCK17}, formalized in (an emulation of a two-level type theory in) the Agda proof assistant~\footnote{\url{https://github.com/nicolaikraus/HoTT-Agda/blob/master/nicolai/SemiSimp/SStypes.agda}}, also rely on the presentation of the semi-simplicial category from increasing injective functions over finite ordinals.

The indexed definition of a presheaf over a direct category is technically more involved than the presheaf definition as it requires hard-wiring in the structure the dependencies between elements of the sets of the presheaf, including the coherence conditions between these dependencies, such as taking the $i$-th face of the $j$-th face of a $n$-simplex is the same as taking the $(j-1)$-th face of the $i$-th face (when $j>i$). On the other side, defining a concrete presheaf in indexed form only requires to provide the family since the responsibility of defining maps and showing the coherence conditions is already accounted for in the definition, what is not the case in the presheaf presentation.

\subsection*{Reynolds' parametricity and its unary and binary variants}
In the context of functional programming, Reynolds' parametricity~(\citeyear{reynolds83}) interprets types as relations characterizing the observational behavior of programs of this type. More generally, correspondences, in the sense of ``relevant'' relation, that is of a family over a product of types, ca Parametricity can be iterated, and it has been noted that iterated Reynolds' parametricity has the structure of a cubical set~\citep{johann17,altenkirch15,moulin16,moeneclaey21,moeneclaey22phd}. We obtain a \emph{unary} variant of Reynolds' \emph{binary} parametricity by using families instead of relations or graphs, in which case, we obtain a form of realizability~\citep{bernardy12,lasson12,moulin16}. It has then been noted that iterated unary parametricity has the structure of an augmented simplicial set~\footnote{Private communication with Hugo Moeneclaey and Thorsten Altenkirch}.

\subsection*{Iterated-parametricity-based construction}
We exploit the connection between iterated unary parametricity and augmented simplicial sets, and between iterated binary parametricity and cubical sets to present a uniform construction of both augmented semi-simplicial and semi-cubical sets, generated by iterating the parametricity translation for type judgements. In contrast to the standard presheaf presentation of augmented semi-simplicial and semi-cubical sets~\citep{fri08,grandis03,buchholtz17}, our uniform construction, which we call \emph{$\nu$-sets}, is \emph{indexed}. That is, instead of having the set of augmented semi-simplices or semi-cubes in dimension $n+1$ \emph{fibered} over the set of augmented semi-simplices or semi-cubes in dimension $n$, we consider families of augmented semi-simplices and semi-cubes indexed over their faces.

\subsection*{Contribution}
Our work:
\begin{itemize}
  \item characterizes, will full details, how to technically define parametricity iterated in arbitrary dimensions
  \item provides a new example of indexed constructions of semi-simplicial and semi-cubical sets~\citep{voevodsky12,herbelin15,part15}, useful in the direction of better understanding the technicality of coherence issues in defining ``semi-simplicial types''~\citep{part15,shulman14,altenkirch16,kraus2017space}, as well as the dialectic between reasoning in Extensional Type Theory and Intensional Type Theory (as will be discussed in Section~\ref{sec:ett})
  \item makes a step in the direction of the program initiated in \cite{altenkirch15} to develop parametricity-based models of parametric type theory~\citep{bernardy15,nuyts17,cavallo19} and cubical type theory~\citep{bezem13,cohen16,angiuli21} which are closer to the syntax of type theory and thus more liable to reflect definitional properties of type theory than presheaf-based cubical sets would do; compare, for example, the loss of definitional properties when interpreting ``indexed'' dependent types of type theory as ``fibrations'' in models such as locally cartesian closed categories~(\citeyear{curien14}).
\end{itemize}
\subsection*{Outline}
The outline of the paper is as follows. We recall the definition of the augmented semi-simplicial and semi-cubical categories using a combinatorial presentation, that directly leads to our generalization to $\nu$-sets, in \ref{sec:nu}. Before spelling out the details of our formalization in type-theoretic language in \ref{sec:ett} and \ref{sec:itt}, we give intuitions in \ref{sec:intuition}. We end with some of the finer details of our mechanization in \ref{sec:mech}.

@@

\section{Semi-simplicial and semi-cubical sets\label{sec:nu}}
In this section, we generalize semi-simplicial and semi-cubical sets to, what we call, $\nu$-sets, subsuming the earlier definitions. However, we also explicitly work out the cases of semi-simplicial and semi-cubical sets, in a way that is similar to a previous art~(\citeyear{aitchison86}).

@@

\section{Preliminaries}
Martin-Löf's Type theory~[\cite{martinlof75,martinlof84}] is a logical formalism based on the notion of a \emph{type} rather than that of a \emph{set}. It can be seen as a foundation of mathematics alternative to set theory and is the core of several tools for the formalization of mathematics such as Agda~\cite{agda23}, Coq~\cite{coq22} or Lean~\cite{lean15}. In type theory, propositions are types and proofs are programs. A particularity of type theory is also that types and programs, hence propositions and proofs also, are considered modulo an equational theory called definitional equality.

@@

The need for transport along a proof of commutation of $\restrf$ in the definition of $\restrl$ is made explicit, where the arrow over $\cohf$ indicates the direction of rewrite. The proof of $\cohf$ requires making explicit several rewritings which were invisible in extensional type theory. The commutation of $\restrl$ lives in a type referring to $\cohf$, so we need a transport along the commutation of $\restrf$ in the statement of $\cohl$. The proof of $\cohl$ is the most involved proof in the construction. This is where the higher-dimensional coherence condition is needed. The exact formulation of this coherence condition is as follows.
\renewcommandx{\restrf}[7][1,2,3,4,5,6,7]{\restr{frame}[][#2][#3][#4][#5][#6][#7]}
\renewcommandx{\cohf}[9][1,2,3,4,5,6,7,8,9]{\coh{frame}[][#2][#3][#4][#5][#6,#7][#8][#9]}
\renewcommandx{\cohl}[9][1,2,3,4,5,6,7,8,9]{\coh{layer}[][#2][#3][#4][#5][#6,#7][#8][#9]}
\renewcommandx{\cohc}[9][1,2,3,4,5,6,7,8,9]{\coh{painting}[][#2][#3][#4][#5][#6,#7][#8][#9]}
\begin{align*}
  \cohf[m][\omega][\theta][r][p][n][p][][d = {\restrf[m][\epsilon][q+2][n][p][][d = \d]}]\;\bullet     \\
  \ap \restrf[m][\omega][r][n][p][][]\;(\cohf[m][\epsilon][\theta][q+1][p][n][p][][d = \d])\;\bullet   \\
  \cohf[m][\epsilon][\omega][q][r][n][p][][d = {\restrf[m][\theta][p][n][p][][d = \d]}] =              \\
  \ap \restrf[m][\theta][p][n][p][][]\;(\cohf[m][\epsilon][\omega][q+1][r+1][n][p][][d = \d])\;\bullet \\
  \cohf[m][\epsilon][\theta][q][p][n][p][][d = {\restrf[m][\omega][r+1][n][p][][d = \d]}]\;\bullet     \\
  \ap \restrf[m][\epsilon][q][n][p][]\;(\cohf[m][\omega][\theta][r][p][n][p][][d = \d])
\end{align*}
where $\mathsf{ap}$ applies a function on two sides of an equality, and $\bullet$ is transitivity of equality. This property of equality proofs holds in \U, and since our formalization is done in \U, the term is trivially discharged.
